---
format: gfm
bibliography: vignettes/references.bib
execute:
  collapse: true
  comment: "#>"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-"
)
set.seed(1234)
```

# badcontrols

<!-- badges: start -->
[![Version](https://img.shields.io/badge/version-`r packageVersion("badcontrols")`-blue.svg)](https://github.com/hugosantanna/badcontrols)
[![R](https://img.shields.io/badge/R-%3E%3D%204.1-blue.svg)](https://cran.r-project.org/)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-green.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

## Overview

A common tension in empirical work involves covariates that are affected by the treatment (aka "bad controls"), where there is an argument for (at least in some sense) trying to control for them, but there are also issues that arise when controlling for them directly.  In @caetano-callaway-payne-santanna-2026, in the context of difference-in-differences identification strategies, we provide two approaches to dealing with bad controls that respect the bad control being affected by the treatment while playing a genuine role as a covariate too.  The `badcontrols` package implements these approaches.

## Approaches

Our paper considers two alternative approaches to identifying the untreated
potential version of the bad control:

1. **Condition on the pre-treatment value of the bad control.** This approach leads to a version of @callaway-santanna-2021 that involves conditioning on the pre-treatment value of the bad control.  It is also simple and can be implemented with standard tools.

2. **Assume covariate unconfoundedness for the bad control.** This generalizes
   the first approach by modeling the bad control's untreated evolution using
   additional covariates.  This approach leads to a more complicated estimator, but the assumptions may be more credible in some applications.

For both approaches, `badcontrols` provides imputation, parametric doubly
robust, and machine-learning estimators. The main function is `didbc()`,
which follows a syntax similar to `did::att_gt()` and
`ptetools::pte_default()`.

## Additional resources

* Paper: @caetano-callaway-payne-santanna-2026.

* [Conceptual vignette](https://hsantanna.org/badcontrols/articles/bad-controls-conceptual.html)

* [Application/Coding vignette](https://hsantanna.org/badcontrols/articles/bad-controls-coding.html)

## Installation

The development version can be installed from GitHub:

```r
install.packages("remotes")
remotes::install_github("hugosantanna/badcontrols")
```

## Basic example

The package includes a simulation with a treatment-affected covariate `X` and exogenous covariate `Z`.  The demo here is based on the covariate unconfoundedness assumption (New Approach 2 above), where we assume that unconfoundedness holds for the bad control after conditioning on the exogenous covariate `Z`, the pre-treatment value of the bad control, and the lagged outcome.  The data that we generate below is a panel with 2000 units and 4 periods.

```{r basic-example}
#| warning: false
library(badcontrols)

sim <- simulate_bad_controls(n = 2000, T_max = 4)

head(sim$data)

res <- didbc(
  yname = "Y",
  gname = "G",
  tname = "period",
  idname = "id",
  data = sim$data,
  bad_control_formula = ~X,
  xformula = ~Z,
  bad_control_cov_formula = ~Y,
  est_method = "dr_ml",
  nuisance_method = "parametric",
  bstrap = FALSE
)

summary(res)
```

The same interface can be used with `est_method = "imputation"` or with
`nuisance_method = "ml"` for cross-fitted machine-learning nuisance
estimates.

## Citation

To cite the paper underlying this package:

> Caetano, C., Callaway, B., Payne, S., and Sant'Anna, H. (2026). "Difference-in-Differences with Bad Controls." arXiv preprint arXiv:2608.03881. https://arxiv.org/abs/2608.03881

```
@article{caetano2026badcontrols,
  title   = {Difference-in-Differences with Bad Controls},
  author  = {Caetano, Carolina and Callaway, Brantly and Payne, Stroud and Sant'Anna, Hugo},
  journal = {arXiv preprint arXiv:2608.03881},
  year    = {2026},
  url     = {https://arxiv.org/abs/2608.03881}
}
```

To cite the `badcontrols` package itself, run `citation("badcontrols")` in R.

## License

GPL (>= 3)

## References
