---
title: "Difference-in-Differences with Bad Controls: A Coding Example"
format: html
vignette: >
  %\VignetteIndexEntry{Difference-in-Differences with Bad Controls: A Coding Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
#| label: setup
#| warning: false
#| message: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the main workflow using simulated staggered-treatment data.
The data-generating process is the linear DGP 1 described in the paper. `X` is
the time-varying bad control, `Y` is the outcome, and `Z` and `W` are
time-invariant covariates.

## Simulated data

```{r}
#| label: simulate-data
#| warning: false
#| message: false

library(badcontrols)
library(ggplot2)

set.seed(123)
sim <- simulate_bad_controls(
  n = 2000,
  T_max = 4,
  groups = 2:4,
  dgp = "dgp1",
  beta_drift = 0
)

head(sim$data)
```

In DGP 1, the untreated evolution of the bad control depends on its lag, `W`,
and `Z`. This determines the covariates used in the bad-control event study
below.

## Event study for the bad control

We first estimate an event study for `X` itself. Because `X` is the outcome in
this exercise, we set `d_outcome = FALSE`: the estimator works with the level
of `X` rather than a change in `X`. The specification includes the lagged bad
control, `W`, and `Z`, matching the DGP 1 evolution equation.

```{r}
#| label: bad-control-event-study
#| warning: false
#| message: false

x_event_study <- ptetools::pte_default(
  yname = "X",
  gname = "G",
  tname = "period",
  idname = "id",
  data = sim$data,
  xformula = ~X + W + Z,
  d_outcome = FALSE,
  d_covs_formula = ~-1,
  est_method = "reg",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE,
  cband = FALSE
)

bad_control_event_study <- data.frame(
  event_time = x_event_study$event_study$egt,
  estimate = x_event_study$event_study$att.egt,
  se = x_event_study$event_study$se.egt
)

bad_control_event_study
```

## Three outcome estimators

We now estimate effects on `Y` in three ways:

1. **Include the bad control:** use the observed post-treatment `X` as a
   covariate.
2. **Drop the bad control:** use only `Z` as a covariate.
3. **Use `badcontrols`:** recover the untreated version of `X` using `W` and
   `Z`, then use it in the outcome comparison.

The first two specifications use `ptetools::pte_default()`. The third uses
`badcontrols::didbc()` with the imputation estimator. All three use the same
staggered-treatment setup and suppress the bootstrap so that the example runs
quickly.

```{r}
#| label: estimate-effects
#| warning: false
#| message: false

pte_args <- list(
  yname = "Y",
  gname = "G",
  tname = "period",
  idname = "id",
  data = sim$data,
  est_method = "reg",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE,
  cband = FALSE
)

include_bad_control <- do.call(
  ptetools::pte_default,
  c(pte_args, list(
    xformula = ~Z + X,
    d_covs_formula = ~X,
    d_outcome = TRUE
  ))
)

drop_bad_control <- do.call(
  ptetools::pte_default,
  c(pte_args, list(
    xformula = ~Z,
    d_covs_formula = ~-1,
    d_outcome = TRUE
  ))
)

badcontrols_imputation <- didbc(
  yname = "Y",
  gname = "G",
  tname = "period",
  idname = "id",
  data = sim$data,
  bad_control_formula = ~X,
  bad_control_cov_formula = ~W,
  xformula = ~Z,
  est_method = "imputation",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE,
  cband = FALSE
)
```

The following plot compares the event-study estimates. Event time `-1` is
normalized to zero by the universal-base-period convention.

```{r}
#| label: plot-outcome-event-studies
#| warning: false
#| message: false

event_study_data <- function(result, estimator) {
  data.frame(
    event_time = result$event_study$egt,
    estimate = result$event_study$att.egt,
    se = result$event_study$se.egt,
    estimator = estimator
  )
}

outcome_event_studies <- rbind(
  event_study_data(include_bad_control, "Include X"),
  event_study_data(drop_bad_control, "Drop X"),
  event_study_data(badcontrols_imputation, "badcontrols imputation")
)

ggplot(outcome_event_studies, aes(event_time, estimate, colour = estimator)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = -1, linetype = "dashed", colour = "grey70") +
  geom_point(position = position_dodge(width = 0.15)) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    width = 0,
    position = position_dodge(width = 0.15),
    na.rm = TRUE
  ) +
  labs(
    x = "Event time",
    y = "Estimated effect on Y",
    colour = "Estimator"
  ) +
  theme_minimal()
```

The conceptual motivation for these comparisons is discussed in the
[conceptual vignette](bad-controls-conceptual.html). The package also provides
parametric doubly robust and machine-learning estimators for the
covariate-unconfoundedness approach.
