---
title: "Mini Application on Job Displacement"
format: html
bibliography: references.bib
header-includes:
  - \newcommand{\E}{\mathbb{E}}
  - \newcommand{\ATT}{\text{ATT}}
  - \newcommand{\independent}{\mathrel{\perp\!\!\!\perp}}
vignette: >
  %\VignetteIndexEntry{Difference-in-Differences with Bad Controls: A Coding Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through the main `badcontrols` workflow using a small scale version of the application in @caetano-callaway-payne-santanna-2026 that considers the effect of job displacement on earnings treating a worker's occupation score as the bad control.

## NLSY data

`nlsy_job_displacement` is a balanced panel of NLSY79 respondents observed
biennially from 1992 to 2002. `log_earnings` is the outcome, `occ_score` is
the bad control (it can change when a respondent changes occupation), and
`group` gives each respondent's displacement year (`0` for never
displaced).

```{r}
#| label: load-data
#| warning: false
#| message: false

library(badcontrols)
library(ptetools)

data(nlsy_job_displacement)
head(nlsy_job_displacement)
table(nlsy_job_displacement$group[!duplicated(nlsy_job_displacement$id)])
```

## Question 1: Is occupation score a bad control?

Before treating `occ_score` as a bad control, it's worth checking directly
whether displacement actually affects it. We can do this with the same
group-time ATT machinery that `didbc()` builds on
(`ptetools::pte_default()`), just using `occ_score` as the outcome instead
of earnings.

```{r}
#| label: check-bad-control
#| warning: false
#| message: false

occ_score_check <- pte_default(
  yname = "occ_score",
  gname = "group",
  tname = "year",
  idname = "id",
  data = nlsy_job_displacement,
  xformula = ~ race + female + educ_max_grade,
  d_outcome = FALSE,
  lagged_outcome_cov = TRUE,
  est_method = "reg",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE
)

summary(occ_score_check)
```

The results here indicate that job displacement reduces the occupation score, especially in the period right after job displacement occurs.

## Estimating the effect of displacement on earnings

`didbc()`'s main arguments mirror `did::att_gt()`/`ptetools::pte_default()`,
plus a few bad-control-specific ones: `bad_control_formula` (the bad
control itself, `occ_score`), `bad_control_cov_formula` (the covariate(s)
used to model its untreated evolution, `W`; here the outcome itself, as in
the paper's application), and `xformula` (the other covariates, `Z`).

Next, we provide estimates of the effect of job displacement on earnings with occupation score treated as a bad control.  We use the imputation version of our estimator (`est_method = "imputation"`).

```{r}
#| label: didbc-imputation
#| warning: false
#| message: false

res_imputation <- didbc(
  yname = "log_earnings",
  gname = "group",
  tname = "year",
  idname = "id",
  data = nlsy_job_displacement,
  bad_control_formula = ~occ_score,
  bad_control_cov_formula = ~log_earnings,
  xformula = ~ race + female + educ_max_grade,
  est_method = "imputation",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE
)

summary(res_imputation)
```

`didbc()` returns event studies alongside the overall ATT, which are plotted below.

```{r}
#| label: plot-event-study
#| warning: false
#| message: false

plot(res_imputation)
```

### Other estimators

The same call works for the doubly robust and machine-learning estimators
(`est_method = "dr_ml"`). Only the relevant arguments change:

```{r}
#| label: other-estimators
#| eval: false

# Doubly robust, parametric (OLS/logit) nuisance functions
didbc(
  yname = "log_earnings",
  gname = "group",
  tname = "year",
  idname = "id",
  data = nlsy_job_displacement,
  bad_control_formula = ~occ_score,
  bad_control_cov_formula = ~log_earnings,
  xformula = ~ race + female + educ_max_grade,
  est_method = "dr_ml",
  nuisance_method = "parametric",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE
)

# Doubly robust, cross-fitted machine-learning (grf) nuisance functions
didbc(
  yname = "log_earnings",
  gname = "group",
  tname = "year",
  idname = "id",
  data = nlsy_job_displacement,
  bad_control_formula = ~occ_score,
  bad_control_cov_formula = ~log_earnings,
  xformula = ~ race + female + educ_max_grade,
  est_method = "dr_ml",
  nuisance_method = "ml",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE
)
```

Finally, instead of covariate unconfoundedness, `didbc()` also supports assuming
*parallel trends for the bad control itself*
(`bad_control_identification_strategy = "did"`).  Since this approach requires an additional linearity condition, it is only available for `est_method = "imputation"`, and still uses
`bad_control_cov_formula` (`W`) if supplied:

```{r}
#| label: didbc-parallel-trends
#| eval: false

didbc(
  yname = "log_earnings",
  gname = "group",
  tname = "year",
  idname = "id",
  data = nlsy_job_displacement,
  bad_control_formula = ~occ_score,
  bad_control_cov_formula = ~log_earnings,
  xformula = ~ race + female + educ_max_grade,
  est_method = "imputation",
  bad_control_identification_strategy = "did",
  control_group = "notyettreated",
  base_period = "universal",
  bstrap = FALSE
)
```



