---
title: "Difference-in-Differences with Bad Controls"
author: "Carolina Caetano, Brantly Callaway, Stroud Payne, Hugo Sant'Anna"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Difference-in-Differences with Bad Controls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## The Bad Controls Problem

In difference-in-differences, researchers often want to include time-varying
covariates in the parallel trends assumption. The problem: if a covariate $X_t$
is measured **after** treatment, it may itself be affected by treatment.
Conditioning on it introduces post-treatment selection bias --- a "bad control."

Three things can go wrong:

1. **Including $X_t$ directly**: introduces bias (absorbs indirect effect $D \to X \to Y$)
2. **Dropping $X_t$ entirely**: may violate parallel trends
3. **Using only $X_{t-1}$**: works under restrictive conditions

This package provides two estimators that correctly handle bad controls:

- **Imputation**: impute the counterfactual $X_t(0)$ for treated units
- **Doubly Robust ML**: semiparametric estimator with cross-fitting

## Simulated Example

We simulate a staggered DiD dataset where:

- Treatment directly affects $Y$ (direct ATT = 0.2)
- Treatment also shifts $X_t$ by $\theta = 0.8$ (bad control!)
- $X_t$ affects $Y_t$ with coefficient $\beta = 1.0$
- True total ATT = $0.2 + 1.0 \times 0.8 = 1.0$

```{r simulate}
library(badcontrols)
set.seed(20240301)

sim <- simulate_bad_controls(n = 2000, T_max = 4)
cat("True ATT:", sim$true_att, "\n")
head(sim$data)
```

## Visualizing the Problem

Treatment causally shifts $X_t$, creating divergent paths between treated and
control groups after treatment onset:

```{r plot, message=FALSE}
library(dplyr)
library(ggplot2)

x_means <- sim$data %>%
  mutate(Group = ifelse(G > 0, "Treated", "Never-Treated")) %>%
  group_by(Group, period) %>%
  summarize(mean_X = mean(X), .groups = "drop")

ggplot(x_means, aes(x = period, y = mean_X, color = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2.5, linetype = "dashed", alpha = 0.5) +
  labs(title = "X diverges after treatment",
       x = "Period", y = "Mean of X", color = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## Estimation

### Imputation Approach

Two-step procedure:

1. Among controls, learn $X_t \sim f(X_{t-1}, W, Z)$
2. For treated, predict counterfactual $X_t(0)$
3. Run DiD using imputed $X_t(0)$

```{r imputation, eval=FALSE}
res_imp <- bc_att_gt(
  yname = "Y", gname = "G", tname = "period", idname = "id",
  data = sim$data,
  bad_control_formula = ~X,
  xformla = ~Z,
  est_method = "imputation"
)
extract_att(res_imp)
```

### Doubly Robust ML

Semiparametric estimator combining outcome regression with propensity score
weighting. Uses random forests (`grf`) with K-fold cross-fitting.

```{r drml, eval=FALSE}
res_ml <- bc_att_gt(
  yname = "Y", gname = "G", tname = "period", idname = "id",
  data = sim$data,
  bad_control_formula = ~X,
  xformla = ~Z,
  est_method = "dr_ml"
)
extract_att(res_ml)
```

## References

Caetano, C., Callaway, B., Payne, S., and Sant'Anna, H. (2024).
"Difference-in-Differences with Bad Controls."
